An Accurate Model for SARS-CoV-2 pooled RT-PCR Test Errors

### 2021-03-24 comment.tex
words and pix

Studies focused on pooling for SARS-CoV-2 are in a consensus that
sample dilution effects \cite{DilutionHIV, GroupDilution} are not a
concern for pools as large as 64 individuals \cite{Lion,
  DorfmanYuvalDor, DilutionCOVID}. Consequently, studies assumed that
the probability of a true-positive (the test's \emph{sensitivity})
does not depend on the number of infected samples in the pool, but
rather on the existence of at least one such sample. Thus, the
probability of a positive result in a pooled test has been (assumed)
identical for a pool with one sample from an infected individual and,
e.g., five such samples. This assumption is common in the group
testing literature \cite{Kim, OptimalDorfmanPool}, as well as in more
specific, COVID-19 focused studies \cite{Simplistic1, Simplistic2}. In
this study we challenge these commonly made assumptions and show how
using a more accurate probabilistic model affects estimation of
false-negative rates for Dorfman pooling.

@@

\subsection{Our model}
Since the essence of the failed common assumption is that
amplification of all samples occurs only once, we assume the opposite:
amplification of viral RNA occurs (or fails) for each sample
independently.  According to \cite{Simplistic1, Simplistic2, Kim,
  OptimalDorfmanPool}, erroneous detection (a false-positive) does not
depend on the number of negative samples in a pool. We incorporate
this assumption in our model with a small modification: We assume that
an erroneous amplification can occur in \emph{any} pool. Specifically,
it is possible that correct amplifications fails and an erroneous one
occurs simultaneously. This assumption is somewhat specific for the
current application of screening for SARS-CoV-2 via RT-PCR. For
example, cross-reactivity with other coronaviruses would have violated
this assumption. However, cross-reactivity was ruled out in
\cite{KitComparison}.

@@

\section{Results}\label{section:results}
\subsection{Scheme false-negative}
We plot $\err{exact}$ for varying prevalence $q$ and sensitivity $\Se$
values, and make the comparison with $\err{common}$. As recommended by
\cite{DorfmanYuvalDor}, we apply different pool sizes $N$, for
different prevalence values. We observe that for a false-positive rate
$\Sp=0.95$ \cite{DorfmanYuvalDor} and a range of reasonable
sensitivity and prevalence values \cite{KitComparison,
  InterpretingCOVID19Test, EstimatingRatesLourenco,
  FalsePositiveEstimate}, an increase of at least $60\%$ in
$\err{exact}$ can be expected (Figure \ref{figy}).

Interestingly, an increase in infection prevalence monotonically
decreases the scheme false-negative rate, as can also be easily seen
from \eqref{eq:sfn}. For the chosen parameter ranges, the increase in
the single test false-negative rates increases the relative error
$\err{exact}$. These effects can be seen in Figure \ref{figy} (left
panel), upon conditioning on pool size. Extending the range for $\Sp$
yields no qualitative differences. We further compare $\err{exact}$ to
$\err{common}$, showing the discrepancy changes as a function of both
prevalence and the single test sensitivity (Figure \ref{figy}, right
panel).
\begin{figure}[H]
  \centering
  \includegraphics[width=18cm]{heatmap_sfn.jpg}
  \caption{Relative increase in Dorfman pooling false-negative rates
    $\err{exact}$. Left: Colors represent $\err{exact}$, the relative
    percentage increase in the scheme false-negative rates relative
    to the single test false-negative rates (eq
    \eqref{eq:erel}). Right: colors represent the difference between
    $\err{common}$ and $\err{exact}$. The disease prevalence $q$, is
    varied on the x-axis, while the test sensitivity is varied on
    the y-axis. Pool size $N$, was chosen according to $q$ as in
    \cite{DorfmanYuvalDor}.}\label{figy}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=18cm]{heatmap_efficiency.jpg}
  \caption{Relative increase in Dorfman pooling efficiency
    $\eff{exact}$. Left: Colors represent $\eff{exact}$, the relative
    percentage increase in the scheme false-negative rates relative
    to the single test false-negative rates (eq
    \eqref{eq:erel}). Right: colors represent the difference between
    $\eff{common}$ and $\eff{exact}$. The disease prevalence $q$, is
    varied on the x-axis, while the test sensitivity is varied on
    the y-axis. Pool size $N$, was chosen according to $q$ as in
    \cite{DorfmanYuvalDor}.}\label{figy}
\end{figure}

@@

we have shown that low values of infection prevalence, or
low values of single test false-negative rates, increase
$\err{exact}$. These results remain qualitatively similar under varying
parameter values, in the observed ranges
\cite{KitComparison,EstimatingRatesKucrika, EstimatingRatesLourenco,
  InterpretingCOVID19Test} (Figure \ref{figy}).

@@

For example, although Dorfman pooling improves testing throughput,

@@

To conclude, pooling is an important technique which can improve
testing throughput in a cost-effective manner. Nevertheless, a
substantial increase in pooling schemes' false-negative rates can be
expected. Such an increase has crucial implications for controlling
the spread of COVID-19.

\bibliographystyle{amsplain}
%\bibliographystyle{ama}
\bibliography{refs}

>>>>

Studies focused on pooling for SARS-CoV-2 are in a consensus that
sample dilution effects \cite{DilutionHIV, GroupDilution} are not a
concern for pools as large as 64 individuals \cite{PoolSize30, Lion,
  DorfmanYuvalDor, DilutionCOVID}. Consequently, studies assumed that
the probability of a true-positive (the test's \emph{sensitivity})
does not depend on the number of infected samples in the pool, but
rather on the existence of at least one such sample. Thus, the
probability of a positive result in a pooled test has been (assumed)
identical for a pool with one sample from an infected individual and,
e.g., five such samples. This assumption is common in the group
testing literature \cite{Kim, OptimalDorfmanPool}, as well as in more
specific, COVID-19 focused studies \cite{Simplistic1, Simplistic2}. In
this study we challenge these commonly made assumptions and show how
using a more accurate probabilistic model affects estimation of
false-negative rates for Dorfman pooling.

@@

\subsection{Our model}\label{subsec:ours}
Since the essence of the failed common assumption is that
amplification of all samples occurs only once, we assume the opposite:
amplification of viral RNA occurs (or fails) for each sample
independently.  According to \cite{Simplistic1, Simplistic2, Kim,
  OptimalDorfmanPool}, erroneous detection (a false-positive) does not
depend on the number of negative samples in a pool. We incorporate
this assumption in our model with a small modification: We assume that
an erroneous amplification can occur in \emph{any} pool. Specifically,
it is possible that correct amplifications fails and an erroneous one
occurs simultaneously. This assumption is somewhat specific for the
current application of screening for SARS-CoV-2 via RT-PCR. For
example, cross-reactivity with other coronaviruses would have violated
this assumption. However, cross-reactivity was ruled out in
\cite{KitComparison}.

@@

\section{Results}\label{section:results}
\subsection{Scheme false-negative}
We plot $\err{exact}$ for varying prevalence $q$ and sensitivity $\Se$
values, and make the comparison with $\err{common}$. As recommended by
\cite{DorfmanYuvalDor}, we apply different pool sizes $N$, for
different prevalence values. We observe that for a false-positive rate
$\Sp=0.95$ \cite{DorfmanYuvalDor} and a range of reasonable
sensitivity and prevalence values \cite{KitComparison,
  InterpretingCOVID19Test, EstimatingRatesLourenco,
  FalsePositiveEstimate}, an increase of at least $60\%$ in
$\err{exact}$ can be expected (Figure \ref{fig1}).

Interestingly, an increase in infection prevalence monotonically
decreases the scheme false-negative rate, as can also be easily seen
from \eqref{eq:sfn}. For the chosen parameter ranges, the increase in
the single test false-negative rates increases the relative error
$\err{exact}$. These effects can be seen in Figure \ref{fig1} (left
panel), upon conditioning on pool size. Extending the range for $\Sp$
yields no qualitative differences. We further compare $\err{exact}$ to
$\err{common}$, showing the discrepancy changes as a function of both
prevalence and the single test sensitivity (Figure \ref{fig1}, right
panel).
\begin{figure}[H]
  \centering
  \includegraphics[width=18cm]{heatmap_sfn.jpg}
  \caption{Relative increase in Dorfman pooling false-negative rates
    $\err{exact}$. Left: Colors represent $\err{exact}$, the relative
    percentage increase in the scheme false-negative rates relative
    to the single test false-negative rates (eq
    \eqref{eq:erel}). Right: colors represent the difference between
    $\err{common}$ and $\err{exact}$. The disease prevalence $q$, is
    varied on the x-axis, while the test sensitivity is varied on
    the y-axis. Pool size $N$, was chosen according to $q$ as in
    \cite{DorfmanYuvalDor}.}\label{fig1}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=18cm]{heatmap_efficiency.jpg}
  \caption{Relative increase in Dorfman pooling efficiency
    $\eff{exact}$. Left: Colors represent $\eff{exact}$, the relative
    percentage increase in the scheme false-negative rates relative
    to the single test false-negative rates (eq
    \eqref{eq:erel}). Right: colors represent the difference between
    $\eff{common}$ and $\eff{exact}$. The disease prevalence $q$, is
    varied on the x-axis, while the test sensitivity is varied on
    the y-axis. Pool size $N$, was chosen according to $q$ as in
    \cite{DorfmanYuvalDor}.}\label{fig2}
\end{figure}

@@

we have shown that low values of infection prevalence, or
low values of single test false-negative rates, increase
$\err{exact}$. These results remain qualitatively similar under varying
parameter values, in the observed ranges
\cite{KitComparison,EstimatingRatesKucrika, EstimatingRatesLourenco,
  InterpretingCOVID19Test} (Figure \ref{fig1}).

@@

%% For example, although Dorfman pooling improves testing throughput,

@@

To conclude, pooling is an important technique which can improve
testing throughput in a cost-effective manner. Nevertheless, care must
be given to pooling schemes' false-negative rates and
efficiency. These are analyzed via a common assumption that we refute.
The implications of refuting the common assumption were presented and
should be considered by practitioners.

%\bibliographystyle{amsplain}
\bibliographystyle{ama}
\bibliography{refs}
